\documentclass{article}
\usepackage{me}
\setlength{\parindent}{0pt}

\title{Introduction\\ 6.854 Notes \#0}
\author{David Karger}

\begin{document}


Welcome.
\begin{itemize}
\item why study this course:
  \begin{itemize}
  \item theoretical elegance, insight on hardness of problems
  \item ability to research in any area of algorithms
  \item ability to recognize problems that arise, apply past techniques,
    and develop new ones
  \item broad sense of ``what is algorithms''.  Not deep---other
    courses follow.
  \item I no longer research algorithms, but the algorithmic perspective often gives new insights on applied problems.
  \end{itemize}
\item Varieties of problems and algorithms
  \begin{itemize}
  \item numerical analysis/linear algebra
  \item number theoretic.  drives cryptography
  \item combinatorial---focus of this course.
    \begin{itemize}
    \item things involving permutations (sorting), graphs (shortest paths)
      and subsets (linear programming).
    \item many {\em optimization problems}---find the best possible
      solution
    \item almost always, finitely many solutions.  brute force always
      works.  we want something better.
    \item combinatorial optimzation: major subarea, but not all we cover
      (vempala course)
    \end{itemize}
  \item aspects of all will arise in others
  \item some problems/algorithms draw from multiple areas---eg comp. geom.
  \end{itemize}
\item Variety of models---sequential, parallel, online, streaming,
  etc.
\item Goals:
  \begin{itemize}
  \item \emph{recognize} an algorithmic opportunity (theory or practice)
  \item know how to \emph{formulate} an algorithmic question
  \item including suitable \emph{computational model}
  \item know where to \emph{seek previous solutions}
  \item be able to \emph{read and understand past work}
  \item know how to \emph{adapt it} to problem variants
  \item some ability to \emph{extend it} to accomplish new things
  \item some ability to \emph{invent new algorithms}
  \end{itemize}
\item course summary sheet
\item Requirements
  \begin{itemize}
  \item psets: wednesdays
  \item project
  \item grading
  \end{itemize}
\item collaboration: 
  \begin{itemize}
  \item have to fully reconstruct the answer yourself.
  \item Cheating is often detected and consequences are severe for both receiver and giver
  \item ethics homework problem on pset 1
  \item if I have to speak to you, we'll have your pset 1 answers.
  \end{itemize}
\item {\bf Experiments}
  \begin{itemize}
  \item live zoom lecture
  \item pre-recorded lecture with realtime Q\&A
  \item pre-recorded lecture with Q\&A after
  \item flipped classroom learning exercises
  \item swap collaborators requirements
    
  \end{itemize}
\item I will teach fast.  Slow me down with questions.
\item Course is time consuming and hard.  
  \begin{itemize}
  \item Collaboration is essential but should not be overused.
  \item A {\bf strong} background in undergraduate algorithms is too
  \end{itemize}
\item Cheating policy on psets.  downgrade if caught.
\end{itemize}

\end{document}
