\documentclass{article}
\usepackage{me}
\setlength{\parindent}{0pt}

\title{Buckets and Van Emde Boas Trees}
\author{David Karger}

\begin{document}
\section{Buckets}

Explot \emph{indirect addresssing} in RAM model.

review shortest path algorithm.

In shortest paths, often have edge lengths small integers (say max
$C$).

Observe heap behavior:
\begin{itemize}
\item heap min increasing (monotone property)
\item max $C$ distinct values 
\item (because don't insert $k+C$ until delete $k$).
\end{itemize}

Idea: lots of things have same value.  Keep in buckets.

How to exploit?  
\begin{itemize}
\item standard heaps of buckets.  $O(m\log C)$
  (slow) or $O(m+n\log C)$ with Fib (messy).
\item Dial's algorithm: $O(m+nC)$.
\item in fact, $O(m+D)$ for max distance $D$
\end{itemize}

space?
\begin{itemize}
\item use array of size $C+1$
\item wrap around
\end{itemize}

Can we improve this?
\begin{itemize}
\item where are we wasting a lot of effort?
\item scanning over big empty regions
\item can we arrange to leap whole regions?
\item some kind of summary structure?
\end{itemize}

2-level buckets.
\begin{itemize}
\item make blocks of size $b$
\item add layer on top: $nC/b$ block summaries
\item in each summary, keep count of items in block
\item insert updates block and summary: $O(1)$
\item ditto decrease key
\item delete min scans summaries, then scans one block
  \begin{itemize}
  \item over whole algorithm, $nC/b$ scanning summaries
  \item also, scan one block per delete min: $b$ work
  \item over $n$ delete mins, work $nb+nC/b$
  \item balance, optimize with $b=\sqrt{C}$
  \end{itemize}
\item result: SP in $O(m+n\sqrt{C})$
\item as before ``space trick'' to keep array sizes to $C$
\item if know $D$ (or binary search for it) make bucket size
  $\sqrt{D/n}$ and get $O(m+\sqrt{nD})$
\item Generalize: 3 tiers?
  \begin{itemize}
  \item blocks, superblocks, and summary
  \item block size $C^{1/3}$
  \item runtime $O(m+nC^{1/3})$
  \item how far can this go?  To $m+n$?  no, because insert cost rises.
  \end{itemize}
\end{itemize}

Tries.
\begin{itemize}
\item depth $k$ tree over arrays of size $\Delta$
\item range $C$ (details of reduction from $nC$ omitted)
\item expansion factor $\Delta=(C+1)^{1/k}$ (power of 2 simplifies)
\item insert: $O(k)$ (also find, delete-non-min, decrease-key)
\item delete-min: $O(k\Delta)=O(kC^{1/k})$ to find next element
\item Shortest paths: $O(km+knC^{1/k})$
\item Balance: $nC^{1/k}=m$ so $C=(m/n)^k$ so $k=\log(C)/\log(m/n)$
\item Runtime: $m\log_{m/n}(C)$
\item Space: $k\Delta=kC^{1/k}=km/n\le (m/n)\log C$ using circular array trick.
\end{itemize}

Problems: space and time

Idea: be lazy!  (Denardo and Fox 1979)
\begin{itemize}
\item don't push items down trie until you must
\item unique block (array) on each level \emph{active}
\item keep other stuff piled up in buckets in each level
\item keep count of items in (buckets of) each level (not counting below)
\item only expand bucket when needed (and update level counts)
\end{itemize}

Operations:
\begin{itemize}
\item Insert
  \begin{itemize}
  \item walk item down tree till stop in (inactive) bucket or bottom
  \item increment level count
  \item $O(k)$ work
  \end{itemize}
\item Decrease key
  \begin{itemize}
  \item Remove from current bucket
  \item put in proper bucket
  \item maybe descend (and revise level counts)
  \item new bucket must be a prior sibling since monotone
    \begin{itemize}
    \item so bucket exists
    \item and no higher levels need be touched
    \end{itemize}
  \end{itemize}
\item amortization:
  \begin{itemize}
  \item items descend once per touch, but never rise, 
  \item (critically relies on monotone property)
  \item so $k$ expansion per item
  \item alternatively, consider ``potential'' as ``total heights of
    items''
  \item inserting at top adds $k$ potential
  \item then all descents are free
  \end{itemize}
\item Delete-min 
  \begin{itemize}
  \item remove item, update bottom level count
  \item advance to new min
    \begin{itemize}
    \item rise to first nonempty level
    \item traverse to first nonempty bucket
    \item expand till find min
    \end{itemize}
  \item time:
    \begin{itemize}
    \item pushdowns are accounted for in insert cost
    \item identifying min happens during pushdowns
    \item rise to nonempty bucket costs less than pushdowns from it
    \item so, cost to scan only one block
    \end{itemize}
  \item cost $C^{1/k}$
  \end{itemize}
\item space to $kC^{1/k}$
\item Times:
  \begin{itemize}
  \item $O(k)$ insert (charge expansions to insert)
  \item $O(1)$ decrease key 
  \item $O(C^{1/k})$ delete-min
  \end{itemize}
\item paths runtime: $O(m+n(k+C^{1/k}))$, choose $k = 2\log C/\log
  \log C$: $O(m+n(\log C)/\log\log C)$
\item great, even if e.g. $C=2^{32}$
\item Further improvement: heap on top (HOT) queues get $O(m+n(\log
  C)^{1/3})$ time.  Cherkassky, Goldberg, and Silverstein.  SODA 97.
\item Implementation experiments---good model for project
\end{itemize}

\section{VEB}


Van Emde Boas, ``Design and Implementation of an efficient priority
queue'' Math Syst. Th. 10 (1977)

Thorup, ``On RAM priority queues'' SODA 1996.

Idea
\begin{itemize}
\item idea: in bucket heaps, problem of finding next empty bucket was
  heap problem.  Recurse!
\item $b$-bit words
\item $\log b$ running times
\item thorup paper improves to $\log\log n$
\item consequence for sorting.
\end{itemize}



Algorithm.  
\begin{itemize}
\item array of buckets but fast lookup on top.
\item queue $Q$ on $b$ bits is struct
  \begin{itemize}
  \item $Q.\min$ is current min, \emph{not} stored recursively
  \item Array $Q.low[]$ of $\sqrt{U}$ queues on low order bits in bucket
  \item $Q.high$, vEB queue on high order bits of elements \emph{other than
    current min in queue}
  \end{itemize}
\item Insert $x$:
  \begin{itemize}
  \item if $x < Q.\min$, swap
  \item now insert $x$ in recursive structs
  \item expand $x=(x_h,x_l)$ high and low half words
  \item If $Q.low[x_h]$ nonempty, then insert $x_l$ in it
  \item else, make new queue holding $x_l$ at $Q.low[x_h]$, and insert
    $x_h$ in $Q.high$
  \item note two inserts, but one to an empty queue, so constant time
  \end{itemize}
\item Delete-min:
  \begin{itemize}
  \item need to replace $Q.\min$
  \item Look in $Q.high.\min$.  if null, queue is empty.  
  \item else, gives first nonempty bucket $x_h$
  \item Delete min from $Q.low[x_h]$ to finish finding $Q.\min$
  \item If results in empty queue, Delete-min from $Q.high$ to remove
    that bucket from consideration
  \item Note two delete mins, but second only happens when first was
    constant time.
  \end{itemize}
\end{itemize}

Problem: space
\begin{itemize}
\item need constant time hash table.  
\item non-trivial complexity theory,
\item but can manage with randomization or slight time loss.
\end{itemize}
\end{document}
